Heart stroke-volume variability in a murine model for heart failure with reduced ejection fraction

### 2023-02-17 Manuscript/manuscript.tex
update

\author{Hugo J. Alves-Figueiredo}

\author{Gerardo de J. García-Rivas}

\affiliation{%
    Tecnológico de Monterrey
}%

@@

There is evidence that increased sympathetic nervous system activity is linked to increased cardiovascular morbidity and mortality \citep{Malpas_2010}. On the other hand, heart rate variability (HRV) analysis has been widely used as a noninvasive assessment tool for autonomic nervous system function \citep{Kiyono_2016}, and results show that decreased and/or abnormal HRV is associated with conditions such as congestive heart failure, diabetic neuropathy, post-cardiac-transplant depression, fatigue severity in chronic fatigue syndrome, and susceptibility to sudden infant dead syndrome \citep{Kamen_1995, Rajendra_Acharya_2006, Zeki_Al_Hazzouri_2014, Escorihuela_2020}.

@@

@@

We conducted the experiments on the control and experimental groups according to the procedures outlined in Section Materials and Methods. We were able to calculate the heart's left ventricle volume as a function of time using the recorded data. This, along with the data on the ventricle pressure that was also recorded, made it possible to reconstruct for each experimental record the so-called P-V diagrams illustrated in Fig. \ref{fig:PVLoops}. Thereafter, we were able to extract time series of physiological interest from these diagrams, such as those of successive cardiac-cycle periods and stroke volumes.

\begin{figure}[h!]
    \includegraphics[width=4in]{PV_Loops.pdf}
    \caption{}
    \label{fig:PVLoops}
\end{figure}

@@

\begin{figure}[h!]
    \begin{tabular}{cc}
        A & B \\
        \includegraphics[width=3in]{Fig01_A.pdf} &
        \includegraphics[width=3in]{Fig01_B.pdf}
    \end{tabular}
    \caption{A) Average cardiac-cycle periods for control and experimental groups. B) Average stroke volumes for control and experimental groups.}
    \label{fig:fig01}
\end{figure}

@@

\begin{figure}[h!]
    \begin{tabular}{cc}
        A & B \\
        \includegraphics[width=3in]{Fig02_A.pdf} &
        \includegraphics[width=3in]{Fig02_B.pdf}
    \end{tabular}
    \caption{}
    \label{fig:fig02}
\end{figure}

@@

\begin{figure}
\includegraphics[width=4in]{model.pdf}
\caption{Schematic representation of the model}
\label{fig:model}
\end{figure}

@@

@@

>>>>

\author{Hugo J. Alves-Figueiredo}
\email{dr.hugo_alves@tec.mx}
\affiliation{%
Cátedra de Cardiología y Medicina Vascular, School of Medicine and Health Sciences, Tecnologico de Monterrey, Monterrey, NL, Mexico
}%

\author{Gerardo García-Rivas}
\email{gdejesus@tec.mx}
\affiliation{%
The Institute for Obesity Research, Tecnologico de Monterrey, Monterrey, NL, Mexico.
}%

@@

There is evidence that increased sympathetic nervous system activity is linked to increased cardiovascular morbidity and mortality \citep{Malpas_2010}. Concomitantly, the sympathetic and vagal nervous systems are key heart-rate regulators, and heart rate variability (HRV) analysis has been widely used as a noninvasive assessment tool for autonomic nervous system function \citep{Kiyono_2016}. Numerous results show that decreased and/or abnormal HRV is associated with conditions such as chronic heart failure, diabetic neuropathy, post-cardiac-transplant depression, fatigue severity in chronic fatigue syndrome, and susceptibility to sudden infant dead syndrome \citep{Kamen_1995, Rajendra_Acharya_2006, Zeki_Al_Hazzouri_2014, Escorihuela_2020}.

@@

\subsection{Animal model of  Heart Failure with reduced ejection Fraction  (HFrEF)}

Mice with induced heart failure were used to study heart rate and stroke volume variability. C57BL/6 male mice weighing 20–30 g were randomly divided into two groups: a control group, and a heart-failure model group. During a week, the mice in the model group were given ad libitum access to drinking water supplemented with 1\% NaCl and 0.01\% of N-nitroL-arginine methyl ester (L-NAME). Following this, a micro-osmotic pump was surgically implanted into the subdermal dorsal area to release angiotensin II (ANGII) at a rate of 0.7 mg/kg/day, based on a prior report \citep{Ruiz_Esparza_2016}. On the other hand, the control group mice were provided plain water and underwent the same surgery but  did not have a pump implanted.

All animal experiments were reviewed and approved by the Institutional Animal Care and Use Committee of the Tecnológico de Monterrey in accordance with the NORMA Oficial Mexicana NOM-062-ZOO-1999.

\subsection{Left-ventricle hemodynamics assessment}

At 28 days post-implantation of the pump, left-ventricle hemodynamics were evaluated in vivo using pressure-volume (PV) analysis with an open-heart configuration and the ADV500 PV measurement system (Transonic Scisense, NY, USA). A 1.2 Fr PV catheter was utilized for this purpose, following the methodology described in a prior publication \citep{Pacher_2008}.

\subsection{Fibrosis and cellular hypertrophy assessment}

After euthanizing the mice, their hearts were extracted and subjected to a semi-quantitative assessment of fibrosis and cellular hypertrophy. This was done by staining the samples with Masson's Trichrome or hemotosilin eosin and capturing microphotographs using an Imager Z.1 Zeiss microscope, equipped with an AxioCam HRm camera. The microphotographs were processed using the AxioVision software. The PV records of mice with hearts that showed no signs of fibrosis and cellular hypertrophy indicative of heart failure were excluded from the study.

\subsection{Statistical analysis}

Statistical analysis was performed to identify significant differences between groups using Student's t-tests with a minimum confidence interval of 95%.

@@

We conducted the experiments on the control and experimental groups according to the procedures outlined in Section Materials and Methods. From the recorded ventricle pressure and volume data we reconstructed for each experimental record the so-called P-V diagrams illustrated in Fig. \ref{fig:PVLoops}. Thereafter, we were able to extract time series of physiological interest from these diagrams, such as those of successive cardiac-cycle periods and stroke volumes.

\begin{figure}[h!]
    \begin{tabular}{cc}
        A & B \\
        \includegraphics[width=3in]{PVLoops_A.pdf} &
        \includegraphics[width=3in]{PVLoops_B.pdf}
    \end{tabular}
    \caption{}
    \label{fig:PVLoops}
\end{figure}

@@

\begin{figure}[h!]
    \begin{tabular}{cc}
        A & B \\
        \includegraphics[width=3in]{Fig01_A.pdf} &
        \includegraphics[width=3in]{Fig01_B.pdf}
    \end{tabular}
    \caption{A) Average cardiac-cycle periods for control (blue) and experimental (orange) groups. B) Average stroke volumes for control (blue) and experimental (orange) groups. Brackets indicates statistically significant difference.}
    \label{fig:fig01}
\end{figure}

@@

\begin{figure}[h!]
    \begin{tabular}{cc}
        A & B \\
        \includegraphics[width=3in]{Fig02_A.pdf} &
        \includegraphics[width=3in]{Fig02_B.pdf}
    \end{tabular}
    \caption{A) Poincairé-plot indices of heart period variability for control (blue) and experimental (orange) groups. B) Poincairé-plot indices of stroke volume variability for control (blue) and experimental (orange) groups. Brackets indicates statistically significant difference.}
    \label{fig:fig02}
\end{figure}

@@

\begin{figure}
\includegraphics[width=4in]{model.pdf}
\caption{Schematic representation of the Frank-Starling model of the cardiovascular system}
\label{fig:model}
\end{figure}

@@

The findings presented in this paper should be approached with caution given the limitations of our methodology. While the absence of HRV in our experimental setup enabled us to establish that stroke volume variability is regulated by distinct mechanisms, it also restricts the generalizability of our results due to its departure from the typical physiological state. To verify our findings, less intrusive methods of evaluating heart hemodynamics are necessary. The simplicity of our mathematical model is an additional constraint in our approach. However, creating and examining a more intricate model requires the estimation of model parameters from experimental data, some of which are not readily accessible in existing literature.

@@

\section*{Data availability statement}

The data that support the findings of this study are openly available at the following URL/DOI:
